---
title: "age_predictive_modeling"
output: pdf_document
date: "2026-06-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries

```{r packages, message=FALSE}
library(data.table)
library(ggplot2)
library(caret)
library(car)
library(randomForest)
library(rpart)
library(tidyverse)
library(pROC)
library(MLmetrics)
library(glmnet)
library(MASS)
library(GGally)
library(earth)
library(Hmisc)

set.seed(123)
```

# Load the Data

```{r}
file_path <- file.choose()
```

# Inspecting the Data

```{r load-data}
nhanes <- fread(file_path)

cat("Dimensions of dataset:", dim(nhanes))
```
# Preprocessing the Data 

```{r format}
# Data preprocessing for Classification
nhanes$age_group <- as.factor(nhanes$age_group)

nhanes$RIAGENDR <- as.factor(nhanes$RIAGENDR)
nhanes$PAQ605 <- as.factor(nhanes$PAQ605)
nhanes$DIQ010 <- as.factor(nhanes$DIQ010)

# Data preprocessing for Regression
reg_data <- reg_data <- nhanes[, .( RIDAGEYR, RIAGENDR, PAQ605, BMXBMI, LBXGLU, 
                                    DIQ010, LBXGLT, LBXIN )]

reg_data$RIAGENDR <- as.factor(reg_data$RIAGENDR)
reg_data$PAQ605 <- as.factor(reg_data$PAQ605)
reg_data$DIQ010 <- as.factor(reg_data$DIQ010)

# Examining Data after Data Type Transformation
str(nhanes)
```
```{r missing-value-check}
# Check for missing values, no imputation needed
colSums(is.na(nhanes))
```
```{r}
# Examining for Degenerate Distributions
nzv <- nearZeroVar(nhanes, names = TRUE)
nzv
table(reg_data$DIQ010)
```


# Exploratory Data Analysis 


```{r, warning=FALSE}
summary(nhanes)
```

## For Regression

```{r, warning=FALSE}
hist.data.frame(reg_data[, .(RIDAGEYR, BMXBMI, LBXGLU, LBXGLT, LBXIN)])
```
```{r}
pairs(reg_data[, .(RIDAGEYR, BMXBMI, LBXGLU, LBXGLT, LBXIN)])
```
```{r boxplots for continuous variables by age}
reg_data$age_bin <- cut(reg_data$RIDAGEYR, breaks = seq(10, 80, by = 10))

par(mfrow = c(1,3))
boxplot(LBXGLU ~ age_bin, data = reg_data,
        xlab = "Age Group (years)", ylab = "Glucose after fasting",
        main = "Fasting Glucose by Age Group")

boxplot(LBXIN ~ age_bin, data = reg_data,
        xlab = "Age Group (years)", ylab = "Blood Insulin Level",
        main = "Insulin Level by Age Group")

boxplot(BMXBMI ~ age_bin, data = reg_data,
        xlab = "Age Group (years)", ylab = "BMI",
        main = "BMI by Age Group")

```
```{r boxplot for categorical predictors}
par(mfrow = c(1,3))

boxplot(RIDAGEYR ~ DIQ010, data = reg_data, subset = DIQ010 != 7 & DIQ010 != 9,
        xlab = "Diabetes Status (1 = Yes, 2 = No, 3 = Borderline)", ylab = "Age (years)",
        main = "Age by Diabetes Status")

boxplot(RIDAGEYR ~ PAQ605, data = reg_data, subset = PAQ605 != 7,
        xlab = "Physical Activity (1 = Yes, 2 = No)", ylab = "Age (years)",
        main = "Age by Physical Activity Status")

boxplot(RIDAGEYR ~ RIAGENDR, data = reg_data,
        xlab = "Gender (1 = Male, 2 = Female)", ylab = "Age (years)",
        main = "Age by Gender")
```

## For Classification

```{r distribution}
table(nhanes$age_group)

ggplot(nhanes, aes(x = age_group, fill = age_group)) +
  geom_bar() +
  scale_fill_manual(values = c("Adult" = "#4A7C8C", "Senior" = "#C97B63")) +
  labs(
    title = "Distribution of Age Group",
    x = "Age Group",
    y = "Count"
  ) +
  theme(legend.position = "none")
```
```{r bmi-box-plot}
ggplot(nhanes, aes(x = age_group, y = BMXBMI)) +
  geom_boxplot() +
  labs(
    title = "BMI by Age Group",
    x = "Age Group",
    y = "Body Mass Index"
  )
```
```{r fasting-glucose-boxplot}
ggplot(nhanes, aes(x = age_group, y = LBXGLU)) +
  geom_boxplot() +
  labs(
    title = "Fasting Glucose by Age Group",
    x = "Age Group",
    y = "Fasting Glucose"
  )
```
```{r oral-glucose-boxplot}
ggplot(nhanes, aes(x = age_group, y = LBXGLT)) +
  geom_boxplot() +
  labs(
    title = "Oral Glucose Tolerance by Age Group",
    x = "Age Group",
    y = "Oral Glucose Tolerance"
  )
```
```{r insulin-level-box-plot}
ggplot(nhanes, aes(x = age_group, y = LBXIN)) +
  geom_boxplot() +
  labs(
    title = "Insulin Level by Age Group",
    x = "Age Group",
    y = "Blood Insulin Level"
  )
```
```{r age-gender-distribution}
ggplot(nhanes, aes(x = RIAGENDR, fill = age_group)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Age Group by Gender",
    x = "Gender: 1 = Male, 2 = Female",
    y = "Count",
    fill = "Age Group"
  )
```
```{r age-physical-distribution}
ggplot(nhanes, aes(x = PAQ605, fill = age_group)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Age Group by Physical Activity",
    x = "Physical Activity: 1 = Yes, 2 = No",
    y = "Count",
    fill = "Age Group"
  )
```
```{r correlation-matrix}
numeric_vars <- c("BMXBMI", "LBXGLU", "LBXGLT", "LBXIN")

cor_matrix <- cor(nhanes[, ..numeric_vars])

round(cor_matrix, 2)
```
```{r correlation-matrix}
cor_long <- as.data.frame(as.table(cor_matrix))
cor_long$label <- round(cor_long$Freq, 2)

ggplot(cor_long, aes(x = Var1, y = Var2, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = label), size = 4) +
  labs(
    title = "Correlation Heatmap of Numeric Predictors",
    x = "Variable",
    y = "Variable",
    fill = "Correlation"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
# Modeling

## For Regression

## For Classification

```{r vif}
vif_model <- lm(
  BMXBMI ~ LBXGLU + LBXGLT + LBXIN,
  data = nhanes
)

vif_values <- vif(vif_model)

vif_table <- data.frame(
  variable = names(vif_values),
  VIF = as.numeric(vif_values)
)

vif_table
```
```{r BMI-fasting-scatterplot}
ggplot(nhanes, aes(x = BMXBMI, y = LBXGLU, color = age_group)) +
  geom_point(alpha = 0.4) +
  labs(
    title = "BMI vs Fasting Glucose by Age Group",
    x = "BMI",
    y = "Fasting Glucose",
    color = "Age Group"
  )
```
# Splitting the Data

```{r predictor-target-variable}
# SEQN is an ID and excluded; RIDAGEYR is excluded as a predictor 
# since age_group is derived from it, which would cause data leakage.

# model data for age_group as target
model_data <- nhanes[, .(
  age_group,
  RIAGENDR,
  PAQ605,
  BMXBMI,
  LBXGLU,
  DIQ010,
  LBXGLT,
  LBXIN
)]

# model data for age as target
reg_model <- reg_data[, !"age_bin"]
```


```{r parition-data}
# partitioning data before addressing the imbalance for age_group as target variable
train_index <- createDataPartition(
  model_data$age_group,
  p = 0.8,
  list = FALSE
)

# partitioning data before addressing the imbalance for age as target variable
train_index_reg <- createDataPartition(
  reg_model$RIDAGEYR,
  p = 0.8,
  list = FALSE
)

# target age
train_reg <- reg_model[train_index_reg, ]
test_reg <- reg_model[-train_index_reg, ]

# target age_group
train <- model_data[train_index, ]
test <- model_data[-train_index, ]

dim(train)
dim(test)

table(train$age_group)
table(test$age_group)

```
# Modeling

## Baseline Model for  Linear Regression
```{r linear model age, warning=FALSE}
ctrl_reg <- trainControl(method = "cv", number = 5)

set.seed(123)
lm_model <- train(
  RIDAGEYR ~ .,
  data = train_reg,
  method = "lm",
  trControl = ctrl_reg
)

summary(lm_model)
```
```{r linear regression predict}
lm_pred <- predict(lm_model, newdata = test_reg)
```

## Basline Model for Logistic Regression
```{r logistic-regression-model age_group}
log_model <- glm(
  age_group ~ RIAGENDR + PAQ605 + BMXBMI + LBXGLU +
    DIQ010 + LBXGLT + LBXIN,
  data = train,
  family = binomial
)

summary(log_model)
```
```{r logistic-regression-predictions age_group}
levels(train$age_group)

log_prob <- predict(log_model, newdata = test, type = "response")

log_pred <- ifelse(log_prob > 0.5, "Senior", "Adult")

log_pred <- factor(log_pred, levels = levels(test$age_group))
```
## Baseline Model for Elastic Net
```{r Elastic Net}
set.seed(123)
glmnet_model_reg <- train(
  RIDAGEYR ~ .,
  data = train_reg,
  method = "glmnet",
  trControl = ctrl_reg,
  tuneGrid = expand.grid(
    alpha = seq(0, 1, 0.2),
    lambda = seq(0.01, 0.2, length = 10)
  )
)

glmnet_model_reg
```
```{r prediction for Elastic Net}
glmnet_pred <- predict(glmnet_model_reg, newdata = test_reg)
```


## Baseline Model for Decisiong Tree with Categorical Target Variable

```{r decision-tree-model age_group}
tree_control <- trainControl(method = "cv", number = 5)

tree_model <- train(
  age_group ~ RIAGENDR + PAQ605 + BMXBMI + LBXGLU +
    DIQ010 + LBXGLT + LBXIN,
  data = train,
  method = "rpart",
  trControl = tree_control,
  tuneLength = 10
)

tree_model
```

## Baseline Model for MARS with continuous Target Variable
```{r}
set.seed(123)
mars_model_reg <- train(
  RIDAGEYR ~ .,
  data = train_reg,
  method = "earth",
  trControl = ctrl_reg,
  tuneGrid = expand.grid(
    degree = 1:2,
    nprune = seq(2, 20, by = 2)
  )
)

mars_model_reg
```
```{r prediction for MARS}
mars_pred <- predict(mars_model_reg, newdata = test_reg)
```



## Baseline Model for Random Forest with Categorical Target Variable

```{r random-forest-model age_group}
rf_control <- trainControl(method = "cv", number = 5)

rf_model <- train(
  age_group ~ RIAGENDR + PAQ605 + BMXBMI + LBXGLU +
    DIQ010 + LBXGLT + LBXIN,
  data = train,
  method = "rf",
  trControl = rf_control,
  tuneLength = 5,
  ntree = 200
)

rf_model
```
```{r original-model-predictions age_group}

log_prob <- predict(log_model, newdata = test, type = "response")
log_pred <- ifelse(log_prob > 0.5, "Senior", "Adult")
log_pred <- factor(log_pred, levels = levels(test$age_group))

tree_pred <- predict(tree_model, newdata = test)

rf_pred <- predict(rf_model, newdata = test)
```

```{r original-model-evaluation-metrics age-group}

log_cm <- confusionMatrix(log_pred, test$age_group)
tree_cm <- confusionMatrix(tree_pred, test$age_group)
rf_cm <- confusionMatrix(rf_pred, test$age_group)

#log_cm
#tree_cm
#rf_cm
```

# Model Results

## Model results for Binary Categorical Target Variable before Imbalance

```{r}
reg_results <- data.frame(
  Model = c("Linear Regression", "Elastic Net", "MARS"),
  RMSE = c(
    RMSE(lm_pred, test_reg$RIDAGEYR),
    RMSE(glmnet_pred, test_reg$RIDAGEYR),
    RMSE(mars_pred, test_reg$RIDAGEYR)
  ),
  MAE = c(
    MAE(lm_pred, test_reg$RIDAGEYR),
    MAE(glmnet_pred, test_reg$RIDAGEYR),
    MAE(mars_pred, test_reg$RIDAGEYR)
  ),
  Rsquared = c(
    R2_Score(lm_pred, test_reg$RIDAGEYR),
    R2_Score(glmnet_pred, test_reg$RIDAGEYR),
    R2_Score(mars_pred, test_reg$RIDAGEYR)
  )
)

reg_results
```

## Model results from our Binary Categorical Target Variable
```{r original-model-performance age_group}

model_results <- data.frame(
  Model = c("Logistic Regression", "Decision Tree", "Random Forest"),
  Accuracy = c(
    log_cm$overall["Accuracy"],
    tree_cm$overall["Accuracy"],
    rf_cm$overall["Accuracy"]
  ),
  Kappa = c(
    log_cm$overall["Kappa"],
    tree_cm$overall["Kappa"],
    rf_cm$overall["Kappa"]
  ),
  Sensitivity = c(
    log_cm$byClass["Sensitivity"],
    tree_cm$byClass["Sensitivity"],
    rf_cm$byClass["Sensitivity"]
  ),
  Specificity = c(
    log_cm$byClass["Specificity"],
    tree_cm$byClass["Specificity"],
    rf_cm$byClass["Specificity"]
  )
)

model_results
```
```{r senior-focused-confusion-matrix age_group}
log_cm_senior <- confusionMatrix(log_pred, test$age_group, positive = "Senior")
tree_cm_senior <- confusionMatrix(tree_pred, test$age_group, positive = "Senior")
rf_cm_senior <- confusionMatrix(rf_pred, test$age_group, positive = "Senior")
```

```{r senior-confusion-matrix age_group}
# 5. Senior-focused comparison table

senior_model_results <- data.frame(
  Model = c("Logistic Regression", "Decision Tree", "Random Forest"),
  Accuracy = c(
    log_cm_senior$overall["Accuracy"],
    tree_cm_senior$overall["Accuracy"],
    rf_cm_senior$overall["Accuracy"]
  ),
  Kappa = c(
    log_cm_senior$overall["Kappa"],
    tree_cm_senior$overall["Kappa"],
    rf_cm_senior$overall["Kappa"]
  ),
  Senior_Sensitivity = c(
    log_cm_senior$byClass["Sensitivity"],
    tree_cm_senior$byClass["Sensitivity"],
    rf_cm_senior$byClass["Sensitivity"]
  ),
  Adult_Specificity = c(
    log_cm_senior$byClass["Specificity"],
    tree_cm_senior$byClass["Specificity"],
    rf_cm_senior$byClass["Specificity"]
  )
)

senior_model_results
```
```{r}
varImp(mars_model_reg)
plot(varImp(mars_model_reg))

plot(test_reg$RIDAGEYR, mars_pred,
     xlab = "Actual Age", ylab = "Predicted Age",
     main = "MARS: Predicted vs. Actual Age")
abline(0, 1, col = "red")
```


```{r original-confusion-matrix age_group}

plot_confusion_matrix <- function(predictions, actual, model_name) {
  
  cm_table <- table(
    Predicted = predictions,
    Actual = actual
  )
  
  cm_df <- as.data.frame(cm_table)
  
  ggplot(cm_df, aes(x = Actual, y = Predicted, fill = Freq)) +
    geom_tile() +
    geom_text(aes(label = Freq), size = 6) +
    labs(
      title = paste(model_name, "Confusion Matrix"),
      x = "Actual Age Group",
      y = "Predicted Age Group",
      fill = "Count"
    )
}

plot_confusion_matrix(log_pred, test$age_group, "Logistic Regression")
plot_confusion_matrix(tree_pred, test$age_group, "Decision Tree")
plot_confusion_matrix(rf_pred, test$age_group, "Random Forest")
```
```{r}
rf_importance <- varImp(rf_model)

rf_importance
plot(rf_importance)
```
```{r balance-training-data}

table(train$age_group)

train_balanced <- upSample(
  x = train[, c("RIAGENDR", "PAQ605", "BMXBMI", "LBXGLU", "DIQ010", "LBXGLT", "LBXIN")],
  y = train$age_group,
  yname = "age_group"
)

table(train_balanced$age_group)
dim(train_balanced)
```
```{r balance-logistic-regression}
log_model_balanced <- glm(
  age_group ~ RIAGENDR + PAQ605 + BMXBMI + LBXGLU +
    DIQ010 + LBXGLT + LBXIN,
  data = train_balanced,
  family = binomial
)

log_prob_balanced <- predict(log_model_balanced, newdata = test, type = "response")

log_pred_balanced <- ifelse(log_prob_balanced > 0.5, "Senior", "Adult")

log_pred_balanced <- factor(
  log_pred_balanced,
  levels = levels(test$age_group)
)
```

```{r balanced-decision-tree}
tree_model_balanced <- train(
  age_group ~ RIAGENDR + PAQ605 + BMXBMI + LBXGLU +
    DIQ010 + LBXGLT + LBXIN,
  data = train_balanced,
  method = "rpart",
  trControl = trainControl(method = "cv", number = 5),
  tuneLength = 10
)

tree_pred_balanced <- predict(tree_model_balanced, newdata = test)

tree_model_balanced
```

```{r balanced-random-forest}
rf_model_balanced <- train(
  age_group ~ RIAGENDR + PAQ605 + BMXBMI + LBXGLU +
    DIQ010 + LBXGLT + LBXIN,
  data = train_balanced,
  method = "rf",
  trControl = trainControl(method = "cv", number = 5),
  tuneLength = 5,
  ntree = 200
)

rf_pred_balanced <- predict(rf_model_balanced, newdata = test)

rf_model_balanced
```

```{r balanced-model-results}
log_cm_balanced <- confusionMatrix(
  log_pred_balanced,
  test$age_group,
  positive = "Senior"
)

tree_cm_balanced <- confusionMatrix(
  tree_pred_balanced,
  test$age_group,
  positive = "Senior"
)

rf_cm_balanced <- confusionMatrix(
  rf_pred_balanced,
  test$age_group,
  positive = "Senior"
)

balanced_results <- data.frame(
  Model = c("Balanced Logistic Regression", "Balanced Decision Tree", "Balanced Random Forest"),
  Accuracy = c(
    log_cm_balanced$overall["Accuracy"],
    tree_cm_balanced$overall["Accuracy"],
    rf_cm_balanced$overall["Accuracy"]
  ),
  Kappa = c(
    log_cm_balanced$overall["Kappa"],
    tree_cm_balanced$overall["Kappa"],
    rf_cm_balanced$overall["Kappa"]
  ),
  Senior_Sensitivity = c(
    log_cm_balanced$byClass["Sensitivity"],
    tree_cm_balanced$byClass["Sensitivity"],
    rf_cm_balanced$byClass["Sensitivity"]
  ),
  Adult_Specificity = c(
    log_cm_balanced$byClass["Specificity"],
    tree_cm_balanced$byClass["Specificity"],
    rf_cm_balanced$byClass["Specificity"]
  )
)

balanced_results
```

```{r confusion matrix for balanced results of random forest}
rf_pred_balanced_releveled <- factor(rf_pred_balanced, levels = c("Senior", "Adult"))
test_age_group_releveled <- factor(test$age_group, levels = c("Senior", "Adult"))

cm_table <- table(
  Predicted = rf_pred_balanced_releveled,
  Actual = test_age_group_releveled
)

cm_df <- as.data.frame(cm_table)

cm_df$Result <- c("True Positive", "False Negative", "False Positive", "True Negative")

ggplot(cm_df, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = paste0(Result, "\n(", Freq, ")")), size = 4.5) +
  labs(
    title = "Balanced Random Forest Confusion Matrix",
    x = "Actual Age Group",
    y = "Predicted Age Group",
    fill = "Count"
  )
```


```{r balanced-original-comparison}
all_model_results <- rbind(
  senior_model_results,
  balanced_results
)

all_model_results
```

```{r roc-auc-curves}
log_prob_balanced_roc <- predict(log_model_balanced, newdata = test, type = "response")
tree_prob_balanced <- predict(tree_model_balanced, newdata = test, type = "prob")[, "Senior"]
rf_prob_balanced <- predict(rf_model_balanced, newdata = test, type = "prob")[, "Senior"]

roc_log <- roc(test$age_group, log_prob_balanced_roc, levels = c("Adult", "Senior"))
roc_tree <- roc(test$age_group, tree_prob_balanced, levels = c("Adult", "Senior"))
roc_rf <- roc(test$age_group, rf_prob_balanced, levels = c("Adult", "Senior"))

plot(roc_log, col = "#4A7C8C",
     legacy.axes = TRUE,
     main = "ROC Curves for Balanced Models",
     xlab = "False Positive Rate", ylab = "True Positive Rate")
lines(roc_tree, col = "#C97B63")
lines(roc_rf, col = "#6A994E")
abline(a = 0, b = 1, lty = 2, col = "gray")

legend("bottomright",
       legend = c(
         paste0("Logistic Regression (AUC = ", round(auc(roc_log), 3), ")"),
         paste0("Decision Tree (AUC = ", round(auc(roc_tree), 3), ")"),
         paste0("Random Forest (AUC = ", round(auc(roc_rf), 3), ")")
       ),
       col = c("#4A7C8C", "#C97B63", "#6A994E"),
       lwd = 2)
```

```{r}
reg_results_by_age <- data.frame(
  actual_age = test_reg$RIDAGEYR,
  predicted_age = as.numeric(mars_pred)
)

reg_results_by_age$age_band <- cut(reg_results_by_age$actual_age, breaks = seq(10, 80, by = 10))

reg_results_by_age |>
  dplyr::group_by(age_band) |>
  dplyr::summarise(RMSE = sqrt(mean((actual_age - predicted_age)^2)))
```

```{r rmse-by-age-band}
rmse_by_band <- reg_results_by_age |>
  dplyr::group_by(age_band) |>
  dplyr::summarise(RMSE = sqrt(mean((actual_age - predicted_age)^2))) |>
  dplyr::filter(!is.na(age_band))

ggplot(rmse_by_band, aes(x = age_band, y = RMSE)) +
  geom_col(fill = "#4A7C8C") +
  geom_text(aes(label = round(RMSE, 1)), vjust = -0.4, size = 3.5) +
  labs(
    title = "MARS Model: RMSE by Age Band",
    x = "Age Band (years)",
    y = "RMSE"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r feature-importance-comparison}
# Logistic Regression - glm object needs different extraction
imp_log_raw <- varImp(log_model_balanced)
imp_log <- data.frame(
  Variable = rownames(imp_log_raw),
  Importance = imp_log_raw$Overall,
  Model = "Logistic Regression"
)

# Decision Tree - caret object
imp_tree_raw <- varImp(tree_model_balanced)$importance
imp_tree <- data.frame(
  Variable = rownames(imp_tree_raw),
  Importance = imp_tree_raw$Overall,
  Model = "Decision Tree"
)

# Random Forest - caret object
imp_rf_raw <- varImp(rf_model_balanced)$importance
imp_rf <- data.frame(
  Variable = rownames(imp_rf_raw),
  Importance = imp_rf_raw$Overall,
  Model = "Random Forest"
)

imp_all <- rbind(imp_log, imp_tree, imp_rf)

ggplot(imp_all, aes(x = reorder(Variable, Importance), y = Importance, fill = Model)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ Model, scales = "free_x") +
  coord_flip() +
  scale_fill_manual(values = c(
    "Logistic Regression" = "#4A7C8C",
    "Decision Tree" = "#C97B63",
    "Random Forest" = "#6A994E"
  )) +
  labs(
    title = "Variable Importance by Model",
    x = "Variable",
    y = "Importance"
  )
```




